Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-05 draft.tex
Created Structure and first chapters

\title{Interactive Visual Analytics of Image-Centric Cohort Study Data}

@@

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous data variables describing the medical condition as well as demographic and lifestyle factors of a subject.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations ('\emph{interactions}').
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\
We propose a Interactive Visual Analytics approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%We propose a Interactive Visual Analytics approach to provide epidemiologists with the tools necessary to analyze this data.
%
It allows for both classical hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead

\section{Medical and Technical Background}
\textbf{Image aquisition}\\
- standardization of MR Scanners

\subsection{Epidemiological Workflow}
Bernhards Paper

\subsection{The Spine Dataset}

\section{Prior and Related Work}
- VMV Paper\\
- Analysis of data in the Rotterdam study\\
- VA in Epidemiology Part of Bernhards Paper
	- Commercial visual Analytics Systems

\section{Interactive Visual Analytics in Cohort Study Data}

@@

\subsection{Feature Localization}
- Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
- Clustering in Image Space yields Groups which can be analyzed using the attribute space

@@

@@

>>>>

\title{Interactive Visual Analysis of Image-Centric Cohort Study Data}

@@

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous data variables describing the medical condition as well as demographic and lifestyle factors of a subject.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations ('\emph{interactions}').
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\
We propose a Interactive Visual Analytics approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%We propose a Interactive Visual Analytics approach to provide epidemiologists with the tools necessary to analyze this data.
%
It allows for both classical hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering. (Remove Clustering reference?)
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims to characterize health and disease by determining risk factors.
%
Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
%
%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
%
Observations made by clinicians in the daily routine are translated into hypothesis.
%
These are used to determine environmental and lifestyle factors are as well as medical attributes which are believed to influence a condition of interest.
%
The data variables necessary are gathered using interviews and clinical examinations.
%
Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
%

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to segment it.
%
Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
\\
We propose a Interactive Visual Analysis approach to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%

%Interactive Visual Analysis (IVA) TODO QUELLE provides techniques which not only show promising potential for combining both 3d visualization with

Our contributions are:
\begin{itemize}
	\item Applying the Interactive Visual Analysis approach to the epidemiological problem domain by characterizing special affordances of this context.
	\item Provide an overview over the workflow for analyzing cohort study data to gain insigth into the large subject spaces.
	\item Provide Visualization Techniques which combine both Information Visualization and 3D Rendering of Organ Shapes as well as combining them with well known epidemiological graphics and key figures.
	\item Implement the presented methods in a Web Framework based on WebGL, D3js and Nodejs as backend.
\end{itemize}

%Statistical correlations derived from analyzing the data itself may be misleading since

%Include Implementation details - web based; how are image data included; what technologies are used?

\section{Medical and Technical Background}
In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
ToDo Define Epidemiological Outcome
%
\subsection{Epidemiological Workflow}
Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis.
%
Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
%
Hypothesis most commonly base on observations made by clinicians in their daily routine.
%
A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
%
Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
%
Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
%

Reproducibility of results is an epidemiological key requirement.
%
Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
%
Graphic data representation is largely used to show results rather than gaining insight.

Bernhards Paper\\
- Missing Data\\
- Follow-Up Questions?\\
- grouping essentiell\\
- problems when analyzing image data\\
	- what are Problems there\\
	- how can shape be included?\\
	- show image data of multiple subjects

\subsection{Epidemiological Data}
Benrhard Paper
- sociodemographic data
- medical data
- pain indicators
- DATA TYPES
	- dichotomous data

\paragraph{Image acquisition}: standardization of MR Scanners

\subsection{The Study of Health in Pomerania (SHIP)}

\section{Prior and Related Work}
- VMV Paper\\
- Analysis of data in the Rotterdam study\\
- VA in Epidemiology Part of Bernhards Paper\\
	- Commercial visual Analytics Systems\\
- GPloms\\
Unterlagen Wissenschaftliche Projekt

\section{Interactive Visual Analytics in Cohort Study Data}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Interaction Overview}
\end{figure}

@@

\subsection{Feature Localization}
- Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
- Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
- Create a Pipeline Overview over different Levels of different IVA Patterns and Stages

@@

\subsection{Implementation}
Diagram of used Technologies

@@

\subsection{The Spine Dataset}
- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
- Input of Epidemiologists goes here!
